### 2025-11-05 sections/1-introduction.tex
removed another half a column

Researchers are collecting data at an unprecedented rate, with datasets in many fields growing exponentially, and outpacing improvements in computing performance predicted by Moore's Law~\cite{kahn2011future}.
Examples include genomic databases~\cite{10.1093/nar/gks1219}, time-series signals~\cite{oshea2018radioml}, and neural embeddings~\cite{2020arXiv200514165B, OpenAI2023GPT4TR, Touvron2023Llama2O, radford2021learning, dosovitskiy2020image}.
This ``Big Data explosion'' has created a need for algorithms that scale efficiently to large datasets.
One such class of algorithms is similarity search algorithms, which enable applications such as recommendation~\cite{annoy} and classification ~\cite{suyanto2022knnclassifier}.
Yet as datasets grow in size and dimensionality, efficient and accurate similarity search becomes increasingly challenging;
even state-of-the-art algorithms exhibit a steep tradeoff between recall and throughput~\cite{malkov2016hnsw, johnson2019billion, annoy, aumuller2020ann}.

Given some measure of similarity between data points, two common similarity search paradigms are $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
The former retrieves the $k$ points closest to a query, while the latter retrieves all points within a similarity threshold $\rho$ of a query.
While prior work has used the term \textit{approximate} search to refer to $\rho$-NN, we use it more narrowly: an {approximate} search algorithm yields imperfect recall, whereas an \textit{exact} algorithm guarantees perfect recall.
$k$-NN search is one of the most widely-used methods in classification and recommendation~\cite{fix1952discriminatory, cover1967nearest}, but na\"{i}ve implementations are prohibitively slow because they scale linearly with dataset size.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for \emph{exact} $k$-NN search.
We benchmark CAKES against FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also evaluate performance on SILVA 18S~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and RadioML~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
To further contextualize results, we also use a synthetic dataset with simple statistical properties.

@@

Several algorithms have been developed to tackle $k$-nearest neighbor search on large datasets, including Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}.

@@

Entropy-scaling algorithms exploit the inherent structure of large datasets, achieving complexity that scales with topological properties, such as metric entropy and fractal dimension, rather than cardinality.
This makes them especially suitable for manifold-constrained data.

In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended the flat clustering from~\cite{yu2015entropy} to a hierarchical approach.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, refines the clustering algorithm from CHESS.

This paper introduces CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust.
We provide a theoretical complexity analysis in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
As a result, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

>>>>

Researchers are collecting data at an unprecedented rate, with datasets in many fields growing exponentially, and outpacing improvements in computing performance predicted by Moore's Law~\cite{kahn2011future}.
Examples include genomic databases~\cite{10.1093/nar/gks1219}, time-series signals~\cite{oshea2018radioml}, and neural embeddings~\cite{2020arXiv200514165B, OpenAI2023GPT4TR, Touvron2023Llama2O, radford2021learning, dosovitskiy2020image}.
This ``Big Data explosion'' has created a need for algorithms that scale efficiently to large datasets.
One such class of algorithms is similarity search;
yet as datasets grow, fast and accurate similarity search becomes increasingly challenging.
Even state-of-the-art algorithms exhibit a steep tradeoff between recall and throughput~\cite{malkov2016hnsw,johnson2019billion,annoy,aumuller2020ann}.

Given some distance function to measure similarity between data points, $k$-nearest neighbors ($k$-NN) search is the task of finding the $k$ points in a dataset that are closest to a given query point.
We use the term \textit{approximate} to refer to algorithms that may not return all true nearest neighbors, while \textit{exact} refers to algorithms that guarantee perfect recall as measured against a brute-force search.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for fast \textit{and} exact search remains~\cite{ukey2023survey}.

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We benchmark these against FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also evaluate performance on SILVA 18S~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and RadioML~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
To further contextualize results, we also use a synthetic dataset with simple statistical properties.

@@

Several algorithms have been developed to tackle $k$-NN search on large datasets, including entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}, Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, and random projection and tree building (ANNOY)~\cite{annoy}.

@@

Entropy-scaling algorithms exploit the inherent manifold structure of large datasets, achieving complexity that scales with topological properties, such as metric entropy and fractal dimension, rather than cardinality.
This makes them especially suitable for manifold-constrained data.

In this paper, we introduce CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust.
These algorithms build upon previous work in CHESS~\cite{ishaq2019clustered}, improving the clustering method and extending to $k$-NN search.
We provide a theoretical analysis of CAKES in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a random uniform distribution.
Given that CAKES's algorithms are intended for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
As a result, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

### 2025-11-05 sections/2-methods.tex
removed another half a column

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, the notion that high($D$)-dimensional data collected from constrained generating phenomena often occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property.
For a point $q \in \textbf{X}$ and radii $r_1 > r_2$,
\begin{equation}
    \text{LFD}(q, r_1, r_2) = \frac{\text{log} \Big( \frac{|B(q, r_1)|}{|B(q, r_2)|} \Big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B(q, r)$ is the set of points in the metric ball of radius $r$ centered at $q$.

% In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
% While we sometimes use Euclidean notions to describe the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
% they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.
% CAKES exploits the low LFD of such datasets to accelerate search.

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases.
When the vast majority of points in the dataset have low ($d \ll D$) LFD, we simply say that the dataset has low LFD.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but (often) globally variable.

@@

We define a \emph{cluster} $C$ as a set of points with a \emph{center} $c$ and \emph{radius} $r$.
The center is the geometric median of the points in the cluster (i.e., the point minimizing the sum of distances to all other points in $C$) and the radius is the maximum distance from the center to any point in $C$.
A cluster with only one point is a $leaf$, otherwise a $parent$ cluster has two $child$ clusters.
Starting from the root cluster $\mathcal{R}$ containing the entire dataset, Alg.~\ref{alg:methods:partition} recursively builds the tree $\mathcal{T}$ in $\mathcal{O}(n \log n)$ time as described in~\cite{ishaq2019clustered}.

@@

        \If{$|C| > 1$ \textbf{and} $C$ satisfies $criteria$}
            \State $S \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
            \State $c \Leftarrow$ geometric median of $S$
            \State $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
            \State $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
            \State $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
            \State $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$
            \State $L \Leftarrow \Call{Partition}{L, criteria}$
            \State $R \Leftarrow \Call{Partition}{R, criteria}$
            % \State $L \Leftarrow Partition(L, criteria)$
            % \State $R \Leftarrow Partition(R, criteria)$
        \EndIf
    \end{algorithmic}
\end{algorithm}

\begin{figure}[t]
    \vskip -0.2in
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \vskip -0.1in
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.2in
\end{figure}
% \FloatBarrier

% \FloatBarrier
\begin{algorithm}[!t]
    \caption{\textsc{TreeSearch}($C$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_C \leq r$}
            \State \textbf{return} $\{C\}$
        \EndIf
        \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
        \State \textbf{return} \Call{TreeSearch}{$L, q, r$} $\cup$ \Call{TreeSearch}{$R, q, r$}
        % \State \textbf{return} tree-search($L, q, r$) $\cup$ tree-search($R, q, r$)
    \end{algorithmic}
\end{algorithm}

@@

Given a query $q$ and integer $k$, $k$-NN search finds the $k$ closest points to $q$ in $\textbf{X}$. In this section, we present three novel algorithms for exact $k$-NN search:
Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
We use $H$ to refer to the data structure containing the nearest neighbors found so far, and $Q$ for the data structure containing the clusters and points that are still in contention for being among the $k$ nearest neighbors.
These algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

We begin using \textsc{TreeSearch} on the root cluster $R$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|X|$.
If the returned set $Q$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $Q$ is non-empty.
Now, so long as $\sum_{C \in Q} |C| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{C \in Q} |C|}} \big)^\frac{1}{\mu}$
where $\mu$ is the harmonic mean of the LFD of the clusters in $Q$ to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low because we need to increase the radius quickly, and small when we have found almost all $k$ points or LFD is large and we need to increase the radius slowly.

Once $\sum_{C \in Q} |C| \geq k$, we perform \textsc{LeafSearch} to find the $k$ nearest neighbors.

@@

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of ranged search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:rnn-search-complexity}
\end{gather}
We now extend this to Repeated $\rho$-NN.

We first need to estimate the number of iterations of \textsc{TreeSearch} needed to find a radius that guarantees at least one neighbor.
Since we expect $q$ to be from the same distribution as $X$, the LFD near $q$ should not differ significantly from the LFDs of clusters near $q$ at the scale of the distance from the query to the nearest neighbor.
Thus, the multiplicative factor described above suggests that in the expected case, we need only one iteration of \textsc{TreeSearch} to find the first cluster, and one more to find enough clusters to guarantee $k$ neighbors.
Since this is a constant factor, Repeated $\rho$-NN has the same complexity for \textsc{TreeSearch} as $\rho$-NN search, i.e.,\,$\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

For \textsc{LeafSearch}, we must estimate $\sum_{C \in Q} |C|$, the total cardinality of those clusters;
since we must examine every point in each such cluster, time complexity of \textsc{LeafSearch} is linear in this quantity.
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is the set of clusters that overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.
Thus, the \textsc{LeafSearch} component of Equation~\ref{eq:methods:rnn-search-complexity} becomes $k \cdot \left( \frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^{\hat{d}}$, where $\hat{d}$ is the harmonic mean of the LFDs of the clusters in $Q$.
It remains to estimate $\rho_k$.

@@

Again using the assumption that $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at a cluster in $Q$, we have that $\hat{d} \approx d$.
Thus, the asymptotic complexity of Repeated $\rho$-NN is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            k \cdot
            \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:repeated-rnn-complexity}
\end{gather}

@@

We begin by letting $Q$ be a list of 3-tuples $(p, \delta^{+}_{p}, m)$, where $p$ is either a cluster or a point, $\delta^{+}_{p}$ is the $\delta^{+}$ of $p$ as illustrated in Figure~\ref{fig:methods:deltas}, and $m$ is the multiplicity of $p$ in $Q$.
During the breadth-first traversal, for every cluster $C$ we encounter, we add $(C, \delta^{+}_{C}, |C| - 1)$ and $(c, \delta_{c}, 1)$ to $Q$, where $c$ is the center of $C$.

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B(q, \tau) \right| \geq k$.
Since this step may require a binary search of $Q$ and reordering $Q$ takes linear time, this version of QuickSelect has $\mathcal{O}\left(|Q| \log |Q|\right)$ time complexity.

Next, we check every item in $Q$, removing any for which $\delta^{-} > \tau$ because they cannot contain (or be) one of the $k$ nearest neighbors.
If an item corresponds to a point, we keep it.
If it corresponds to a leaf cluster, we add all its points to $Q$ with a multiplicity of 1 each.
Otherwise (since it must correspond to a parent cluster), we add to $Q$ the 3-tuples corresponding to its child clusters, thus expanding the search to the next level in $T$.
We continue until the sum of multiplicities in $Q$ is exactly $k$.
These are the $k$ nearest neighbors.

@@

This algorithm (DFS) uses a max-priority queue to track hits, and a min-priority queue to choose the next branch of $\mathcal{T}$ to explore, making it similar to a depth-first traversal.

Let $Q$ be a min-queue of clusters prioritized by $\delta^{-}$ and $H$ be a max-queue (with capacity $k$) of points prioritized by $\delta$.
$Q$ starts containing only $\mathcal{R}$ while $H$ starts empty.
So long as $H$ is not full or the top-priority items from $H$ and $Q$ have $\delta_H \geq \delta_Q$, we repeat following:
\begin{itemize}
    \item While the top-priority cluster is not a leaf, remove it from $Q$ and add its children to $Q$.
    \item Remove the top-priority cluster (a leaf) from $Q$ and add all its points to $H$.
\end{itemize}
This process terminates when $H$ is full and the top-priority items in $H$ and $Q$ are such that $\delta_H < \delta_Q$, i.e.,\,the next closest point to be visited is farther from the query than the $k^{th}$ nearest neighbor found so far.
Thus, $H$ contains the $k$ nearest neighbors to the query.

@@

Due to their similarity, we combine the complexity analyses of both Sieve methods.
For these methods we again use the terminology of \textsc{TreeSearch} and \textsc{LeafSearch}.
Let $d$ be the LFD in the region around $q$, and consider clusters with cardinalities near $k$.
Then the number of such clusters is bounded above by $2d$, where the bound is achieved by having a cluster overlap the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for \textsc{TreeSearch}, these clusters would all come from different branches of the tree, and \textsc{TreeSearch} would look at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity of this component is $T \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For \textsc{LeafSearch}, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the complexity is $L \coloneqq \mathcal{O} \Big( k \cdot \Big( 1 + 2 \cdot \big( \frac{\hat{|C|}}{k} \big) ^ {d^{-1}} \Big)^d \Big)$.

The asymptotic complexity of BFS is dominated by QuickSelect.
Since $Q$ contains the clusters from \textsc{TreeSearch} and the points from \textsc{LeafSearch}, we see that the asymptotic complexity of BFS is
\begin{equation}
    \mathcal{O} \big( (T + L ) \log (T + L ) \big).
    \label{eq:methods:bfs-complexity}
\end{equation}
Similarly, the asymptotic complexity of DFS is dominated by the priority queue operations for $T$ clusters and $L$ points;
it is given by
\begin{equation}
    \mathcal{O} \big( T \log T + L \log k \big).
    \label{eq:methods:dfs-complexity}
\end{equation}

>>>>

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, the notion that high($D$)-dimensional data collected from constrained generating phenomena often occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property.
For a point $q \in \textbf{X}$ and radii $r_1 > r_2$,
\begin{equation}
    \text{LFD}(q, r_1, r_2) = \frac{\text{log} \Big( \frac{|B(q, r_1)|}{|B(q, r_2)|} \Big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B(q, r)$ is the set of points within radius $r$ from $q$.

When the majority of points in the dataset have low ($d \ll D$) LFD, we simply say that the dataset has low LFD.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but (often) globally variable.

@@

We define a \emph{cluster} $C$ as a set of points with a \emph{center} and \emph{radius}.
The center $c$ is the geometric median of the points in the cluster (i.e., the point minimizing the sum of distances to all other points in $C$) and the radius $r$ is the maximum distance from $c$ to any point in $C$.
A cluster with only one point is a $leaf$, otherwise a $parent$ cluster has two $child$ clusters.
Starting from the root cluster $\mathcal{R}$ containing the entire dataset, Alg.~\ref{alg:methods:partition} recursively builds the tree $\mathcal{T}$ in $\mathcal{O}(n \log n)$ time.

\begin{figure}[t]
    \vskip -0.1in
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.05in
\end{figure}

@@

        \If{$|C| > 1$ \textbf{and} $C$ satisfies $criteria$}
        \State $S \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
        \State $c \Leftarrow$ geometric median of $S$
        \State $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \State $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \State $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \State $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$
        \State \Call{Partition}{L, criteria}
        \State \Call{Partition}{R, criteria}
        \EndIf
    \end{algorithmic}
\end{algorithm}

Given a query $q$ and integer $k$, $k$-NN search finds the $k$ closest points to $q$ in $\textbf{X}$. In this section, we present three novel algorithms for exact $k$-NN search:
Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
We use $Q$ for the data structure containing the clusters that are still in contention for being among the $k$ nearest neighbors, and $H$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
These algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

We begin using \textsc{TreeSearch} on the root cluster $R$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|X|$.
If the returned set $Q$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $Q$ is non-empty.
Now, so long as $\sum_{C \in Q} |C| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{C \in Q} |C|}} \big)^\frac{1}{\mu}$
where $\mu$ is the harmonic mean of the LFD of the clusters in $Q$ to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low because we need to increase the radius quickly, and small when we have found almost all $k$ points or LFD is large and we need to increase the radius slowly.
Once $\sum_{C \in Q} |C| \geq k$, we perform \textsc{LeafSearch} to find the $k$ nearest neighbors.

\begin{algorithm}[!t]
    \caption{\textsc{TreeSearch}($C$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_C \leq r$}
            \State \textbf{return} $\{C\}$
        \EndIf
        \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
        \State \textbf{return} \Call{TreeSearch}{$L, q, r$} $\cup$ \Call{TreeSearch}{$R, q, r$}
    \end{algorithmic}
\end{algorithm}

@@

@@

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of ranged search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:rnn-search-complexity}
\end{gather}

To extend this to Repeated $\rho$-NN, we first need to estimate the number of iterations of \textsc{TreeSearch} needed to find a radius that guarantees at least one neighbor.
Since we expect $q$ to be from the same distribution as $X$, the LFD near $q$ should not differ significantly from the LFDs of clusters near $q$ at the scale of the distance from the query to the nearest neighbor.
Thus, the multiplicative factor described above suggests that in the expected case, we need only one iteration of \textsc{TreeSearch} to find the first cluster, and one more to find enough clusters to guarantee $k$ neighbors.
Since this is a constant factor, Repeated $\rho$-NN has the same complexity for \textsc{TreeSearch} as $\rho$-NN search, i.e.,\,$\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

For \textsc{LeafSearch}, we must estimate $\sum_{C \in Q} |C|$;
since we must examine every point in each such cluster, time complexity of \textsc{LeafSearch} is linear in this quantity.
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is the set of clusters that overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.
Thus, the \textsc{LeafSearch} component of Equation~\ref{eq:methods:rnn-search-complexity} becomes $k \cdot \left( \frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^{\hat{d}}$, where $\hat{d}$ is the harmonic mean of LFDs in $Q$.
It remains to estimate $\rho_k$.

@@

Again using the assumption that $q$ is from the same distribution as $X$, we have that $\hat{d} \approx d$.
Thus, the asymptotic complexity of Repeated $\rho$-NN is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            k \cdot
            \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg).
    \label{eq:methods:repeated-rnn-complexity}
\end{gather}

@@

We begin by letting $Q$ be a list of 3-tuples $(C, \delta^{+}, m)$, where $C$ is a cluster, $\delta^{+}$ is as illustrated in Figure~\ref{fig:methods:deltas}, and $m$ is the multiplicity of $C$ in $Q$.
$Q$ starts containing only $(\mathcal{R}, \delta^{+}_{\mathcal{R}}, |\mathcal{R}| - 1)$ while $H$ starts with the center of $\mathcal{R}$.

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B(q, \tau) \right| \geq k$ (including the points in $H$).
Since this step may require a binary search of $Q$ and reordering $Q$ takes linear time, this version of QuickSelect has $\mathcal{O}\left(|Q| \log |Q|\right)$ time complexity.

Next, we check every cluster in $Q$, removing any for which $\delta^{-} > \tau$ because they cannot contain one of the $k$ nearest neighbors.
For each of the remaining clusters, if it is a leaf, we remove it from $Q$ and add all its points to $H$.
Otherwise (since it must be a parent cluster), we remove it from $Q$ and, for each of its child clusters, we compute $\delta^{+}$ and add the corresponding 3-tuple to $Q$ and the child's center to $H$, thus expanding the search to the next level in $T$.
We continue until $Q$ is empty or the farthest point in $H$ is closer to $q$ than the $\delta^{-}$ of every cluster in $Q$.
Thus, $H$ contains the $k$ nearest neighbors to the query.

@@

This algorithm (DFS) greedily descends into the child cluster whose center if closer to the query than that of its sibling.

Let $Q$ be a min-queue of clusters prioritized by $\delta^{-}$ starting with only $\mathcal{R}$ while $H$ starts empty.
So long as $H$ is not full or the top-priority items from $H$ and $Q$ have $\delta_H \geq \delta_Q$, we repeat following:
\begin{itemize}
    \item While the top-priority cluster is not a leaf, remove it from $Q$ and add its children to $Q$.
    \item Remove the top-priority cluster (a leaf) from $Q$ and add all its points to $H$.
\end{itemize}
This process terminates when $H$ is full and the top-priority items in $H$ and $Q$ are such that $\delta_H < \delta_Q$, i.e.,\,the next closest point to be visited is farther from the query than the $k^{th}$ nearest neighbor found so far.
Thus, $H$ contains the $k$ nearest neighbors to the query.

@@

Due to their similarity, we combine the analyses of BFS and DFS.

Let $d$ be the LFD in the region around $q$, and consider clusters with cardinalities near $k$.
Then the number of such clusters is bounded above by $2d$, where the bound is achieved by having a cluster overlap the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for \textsc{TreeSearch}, these clusters would all come from different branches of the tree, and \textsc{TreeSearch} would look at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity of this component is $T \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For \textsc{LeafSearch}, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the complexity is $L \coloneqq \mathcal{O} \Big( k \cdot \Big( 1 + 2 \cdot \big( \frac{\hat{|C|}}{k} \big) ^ {d^{-1}} \Big)^d \Big)$.

The asymptotic complexity of BFS is dominated by our modified QuickSelect algorithm, which accounts for the clusters in $Q$ and the points in $H$, and is log-linear in their sum.
Thus, the asymptotic complexity of BFS is
\begin{equation}
    \mathcal{O} \big( (T + L) \log(T + L) \big).
    \label{eq:methods:bfs-complexity}
\end{equation}
Similarly, the asymptotic complexity of DFS is dominated by the priority queue operations for $T$ clusters and $L$ points;
it is given by
\begin{equation}
    \mathcal{O} \big( T \log T + L \log L \big).
    \label{eq:methods:dfs-complexity}
\end{equation}
